## Comparison with stacked bar charts

Stacked bar charts are commonly used to visualize community composition, with different colors corresponding to taxa.
However, the use of color for categories limits the number of number of taxa displayed to the number of colors that are easily decernable, which is around 10. 
Therefore, rare taxa are often exculded from stackaed barcharts or only coarse taxonomic ranks are displayed. 
This might leave out biolicoally relevant information. 
The code below visulizes the community composition of the same 2 samples from the HMP using both a stacked barchart and heat trees so the two methods can be compared.

### Graph heat trees

```{r single_sample_trees, cache = TRUE}
plot_sample_tree <- function(obj, sample, out_name, min_count = min_read_count, ...) {
  to_use <- obj$data$tax_prop[["700114488"]] >= 0.001 | obj$data$tax_prop[["700114430"]] >= 0.001
  obj %>%
    mutate_obs("tax_prop", sample_prop = obj$data$tax_prop[[sample]]) %>%
    filter_taxa(to_use) %>%
    heat_tree(
      ...,
      title_size = .05,
      node_size = sample, 
      node_color = sample, 
      node_label = ifelse(sample >= 0.01, taxon_names, NA),
      node_label_max = 50, 
      node_size_range = c(.01, .07),
      node_label_size_range = c(.015, .03), 
      node_color_axis_label = "Proportion of reads",
      output_file = result_path(out_name))
}

sample_1 <- plot_sample_tree(hmp_data, "700114488", "sample_1_tree", title = "Sample 1") 
sample_2 <- plot_sample_tree(hmp_data, "700114430", "sample_2_tree", title = "Sample 2") 
```

### Stacked barchart

```{r}
min_count <- 1
bar_data <- filter_taxa(v35_data, (sample_1 >= min_count | sample_2 >= min_count) & name != "") %>%
  calculate_prop("700114482", "sample_1") %>%
  calculate_prop("700114439", "sample_2") %>%
  taxon_data()

bar_data <- bar_data[bar_data$rank == "p", c("name", "sample_1", "sample_2")]
bar_data <- reshape2::melt(bar_data, id.vars = "name")

bar_data <- bar_data[! bar_data$name %in% c("GN02", "SR1"), ]
```

```{r}
library(ggplot2)
bar_plot <- ggplot(bar_data, aes(x = variable, y = value, fill = name)) + 
  geom_bar(stat = "identity") + 
  scale_x_discrete(labels = c("1", "2")) +
  xlab("Sample") + 
  theme_minimal() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_blank(),
        axis.title.y=element_blank(),
        axis.title.x=element_text(size=14),
        axis.text.y=element_blank(),
        axis.text.x=element_text(size=14, margin=margin(b = 6, t = -15)),
        axis.ticks.y=element_blank(),
        legend.title=element_blank(),
        legend.text=element_text(size=12),
        legend.position=c(1,1), 
        legend.justification=c(0, 1), 
        # legend.key.width=unit(1.2, "lines"), 
        plot.margin = unit(c(-0.35, 8, 0, 0.5), "lines"))
ggsave(bar_plot, filename = result_path("bar_chart"), dpi = 350, width = 8, height = 8)
```

### Combine plots

```{r fig.width=7, fig.height=2.61}
library(gridExtra)
library(cowplot)
plot_name <- "barchart_heat_tree_comparison"
combined_plot <- plot_grid(bar_plot, sample_1, sample_2, ncol = 3, nrow = 1, rel_widths = c(1, 2, 2))
save_plot(filename = result_path(plot_name),
          combined_plot, ncol = 3, nrow = 1, base_aspect_ratio = .9)
save_publication_fig(plot_name, figure_number = 2)
print(combined_plot)
```

Note how the two visulization methods can suggest different conclusions regarding how similar the two communities are. 
The stacked barcharts suggest that communities are dominated by Firmicutes, but sample 2 has additional Bacteriodetes and Proteobacteria.
They look different, but not too different.
The heat trees tell a much differnt story. 
Almost all of sample 1 is composed of two genera, Lactobacillus and Gardnerella, which are completly absent in sample 2.
In fact, the samples do not even share any families, so the similarity suggested by the stacked barchart is misleading.
We could have construcated the stacked barchart at the family level to see these differneces, but then there would be too many colors to easily distinguish.

To be fair, I did choose these two samples to prove this point. 
In other cases stacked barcharts might represent data well, but using heat trees instead of stacked barcharts will avoid these kind of misleading graphs. 
