---
output: html_document
bibliography: "bibliography.bibtex"
---

```{r setup, include=FALSE}
source("settings.R")
knitr::opts_chunk$set(prompt = TRUE, comment = NA)
library(dplyr)
library(tibble)
library(readr)
```


This page contains instructions on how to install the dependencies for this workshop without internet access.
If you are doing this, you are probably reading this page from a file provided by the instructor rather than over the internet.
Along with the website files (what you are reading now), there should be a directory called "software" that contains the software to install.
**If you have internet access, use the standard way of installing software described [here](workshop--00--required_software.html)**


## Installing R

For Mac and Windows, use the installer included in the "software/r/mac" or "software/r/windows".
This tutorial does currently not provide a way to install R on Linux offline.

## Installing Rstudio

Use the installers in "software/rstudio/mac, "software/rstudio/windows" or "software/rstudio/linux", depending on your operating system.

## Installing R packages

The following packages are used in this workshop:

```{r include = FALSE}
needed_packages <- c("vegan", "metacoder", "taxa", "ggplot2", "dplyr", "readr", "stringr", "agricolae", "ape", "phyloseq")
```

```{r echo=FALSE}
needed_packages
```

However, these depend on many other packages that must also be installed, so the complete list is:

```{r include = FALSE}
# based on https://www.r-bloggers.com/installing-packages-without-internet/
get_dependencies <- function(packs) {
  dependencyNames <- unlist(
    tools::package_dependencies(packages = packs, 
                                db = available.packages(repos = BiocManager::repositories()), 
                                which = c("Depends", "Imports"),
                                recursive = TRUE))
  packageNames <- union(packs, dependencyNames)
  packageNames
}

all_packages <- get_dependencies(needed_packages)
```

```{r echo=FALSE}
all_packages
```


```{r include=FALSE}
windows_path <- file.path('software', 'packages', 'windows')
osx_path <- file.path('software', 'packages', 'mac_osx_el_capitan')
source_path <- file.path('software', 'packages', 'source')
pkg_info_path <- file.path('software', 'packages', 'package_info.csv')
```

```{r include=FALSE, eval=FALSE}
# This chuck is invisible and not evaluated when this page is built.
# Run manually to update packages included in the "software" folder.

# Make folders to store packages in
dir.create(windows_path, recursive = TRUE)
dir.create(osx_path, recursive = TRUE)
dir.create(source_path, recursive = TRUE)

# Download packages
win_pkg_info <- download.packages(pkgs = all_packages, 
                                  destdir = windows_path,
                                  repos = BiocManager::repositories(),
                                  type = "win.binary",
                                  quiet = TRUE)
osx_pkg_info <- download.packages(pkgs = all_packages, 
                                  destdir = osx_path,
                                  repos = BiocManager::repositories(),
                                  type = "mac.binary.el-capitan",
                                  quiet = TRUE)
source_pkg_info <- download.packages(pkgs = all_packages, 
                                     destdir = source_path,
                                     repos = BiocManager::repositories(),
                                     type = "source",
                                     quiet = TRUE)
colnames(win_pkg_info) <- c('package', 'win_path')
colnames(osx_pkg_info) <- c('package', 'osx_path')
colnames(source_pkg_info) <- c('package', 'source_path')
win_pkg_info <- as_tibble(win_pkg_info)
osx_pkg_info <- as_tibble(osx_pkg_info)
source_pkg_info <- as_tibble(source_pkg_info)
combined_pkg_info <- source_pkg_info %>%
  left_join(win_pkg_info) %>%
  left_join(osx_pkg_info) 
write_csv(combined_pkg_info, path = pkg_info_path)
```

### Windows

```{r eval=FALSE}
setwd('path/to/workshop/folder') # If the workshop folder is an rstudio project, this is not needed
pkg_info <- read.csv(file = 'software/packages/package_info.csv', stringsAsFactors = FALSE)
install.packages(pkg_info$win_path, repos = NULL, type = "win.binary")
```

### macOS and Linux

```{r eval=FALSE}
setwd('path/to/workshop/folder') # If the workshop folder is an rstudio project, this is not needed
pkg_info <- read.csv(file = 'software/packages/package_info.csv', stringsAsFactors = FALSE)
install.packages(pkg_info$source_path, repos = NULL, type = "source")
```

### Mac OSX (El Capitan)

```{r eval=FALSE}
setwd('path/to/workshop/folder') # If the workshop folder is an rstudio project, this is not needed
pkg_info <- read.csv(file = 'software/packages/package_info.csv', stringsAsFactors = FALSE)
install.packages(pkg_info$osx_path, repos = NULL, type = "mac.binary.el-capitan")
```


