---
output: html_document
bibliography: "bibliography.bibtex"
---

```{r setup, include=FALSE}
source("settings.R")
knitr::opts_chunk$set(prompt = TRUE, comment = NA)
library(DT)
```

# Reading data into the `taxmap` format

Getting your data into the proper format one of the more tedious and annoying tasks in bioinformatics.
The `taxa` package supplies generic parsers that can read almost any format of taxonomic data, but they often require some understanding of regular expressions.
The `metacoder` package has easier-to-use functions to parse common formats, but there might not be a function for your particular formats.
This page shows examples for how to parse specific formats, using either the dedicated `metacoder` functions or the more general `taxa` functions.


## Dada2

Unlike most metabarcoding pipelines, dada2 is entirely in R, so the abundance matrix and taxonomy information returned will be stored in a `matrix` in R rather than a .txt file.
Below is an example of such data, taken from the [dada2 tutorial](https://benjjneb.github.io/dada2/tutorial.html):

```{r}
load("cookbook_data/dada2.RData")
DT::datatable(seqtab.nochim)
DT::datatable(taxa)
seq_table = seqtab.nochim
tax_table = taxa
parse_dada2(seq_table, tax_table)
```

