---
output: html_document
bibliography: "bibliography.bibtex"
---

```{r init, echo = FALSE, message = FALSE}
library(knitr)
if (! is.null(current_input())) { # if knitr is being used
  knitr::read_chunk("settings.R")
} else {
  source("settings.R")
}
```

```{r rendering_settings, include=FALSE}
```


# Frequently asked questions

### Error: otu_table slot is empty

This happens when you have `phyloseq` loaded after `metacoder`. 
There is also a `filter_taxa` in `phyloseq`, so the one from the `taxa` package is masked.
Try prefixing `filter_taxa` with `taxa::` or loading `metacoder` after `phyloseq`.

### The labels are too small!

This is a common issue.
For plots with many taxa (> 300 or so) it might not be possible to make a plot where all taxon labels will be readable in a standard publication figure without zooming in.
However, there are a few tricks can help maximize the readability of your labels:

1) Only plot the taxa relevant to the purpose of the figure by filtering using `filter_taxa` before passing the result to `heat_tree`. If you are trying to show the diversity of an entire community, consider filtering out low-abundance or otherwise rare taxa. For example:

```{r eval=FALSE}
my_taxmap %>%
  filter_taxa(taxon_names == "Bacteria", subtaxa = TRUE) %>%
  filter_taxa(taxon_count > 10) %>%
  heat_tree(
    ...
  )
```


2) Selectively suppress labels that do not need to be printed. For example, if you have taxon names like "unknown", they don't really need to be printed, even if you want the node to show in the graph. You can then say that unlabeled taxa are unknown in the figure caption. For example:

```{r eval=FALSE}
dont_print <- c("unknown", "unidentified")
heat_tree(
  ...,
  node_label = ifelse(taxon_names %in% dont_print, "", taxon_names),
  ...
)
```

3) Try changing the layout of the graph. The default layout does not work well for all taxonomies. There are others that use simulated annealing to space out nodes more, which might make more room for labels. You should experiment with the `layout` and `initial_layout` options to see what works best. I like the following settings often:

```{r eval=FALSE}
heat_tree(
  ...,
  layout = "davidson-harel",
  initial_layout = "reingold-tilford",
  ...
)
```


4) Force labels to be a specified size

By default the labels are relative to the size of the nodes. If the nodes are really small, so are the labels. You can force the labels to be a specific size range using the `node_label_size_range` and `edge_label_size_range`. These options take a two-value vector of numbers between 0 and 1 indicating the proportion the the plotted area the text size is. For example, to make text height between 1% and 3% of the plot size:

```{r eval=FALSE}
heat_tree(
  ...,
  node_label_size_range = c(0.01, 0.03),
  ...
)
```

You could also use `node_size_range` and `edge_size_range` to change the size of the labels as a side effect of changing the size of nodes/edges.

5) Shorten labels

Do your labels have taxonomic rank codes or other unneeded text? If so get rid of it using text replacing functions that can use `gloss$add('reqular expressions')` like `sub` and `gsub`. Labels can also be printed over multiple lines, so you might consider replacing some spaces with newlines (`\n`).


6) Split the plot into multiple plots

If you have done all the above and the plot is still too dense, you can turn it into multiple plots of subsets of the data.
Large groups you can give their own tree by filtering the taxonomy using `filter_taxa`:

```{r eval=FALSE}
my_taxmap %>%
  filter_taxa(taxon_names == "Ascomycota", subtaxa = TRUE) %>%
  heat_tree(
    ...
  )

my_taxmap %>%
  filter_taxa(taxon_names == "Basidiomycota", subtaxa = TRUE) %>%
  heat_tree(
    ...
  )
```

If you have a bunch of small groups left over after plotting the big ones, make a single tree of those by removing the big ones plotted elsewhere using the `subtaxa` and `invert` options:

```{r eval=FALSE}
my_taxmap %>%
  filter_taxa(taxon_names %in% c("Ascomycota", "Basidiomycota"), subtaxa = TRUE, invert = TRUE) %>%
  heat_tree(
    ...
  )
```


### How do I get my data into metacoder?

That depends on your data's format. For some formats, there are easy to user "parsers". All of them start with `parse_`, so you can type (after loading `metacoder`) `parse_` and hit tab and see what are available. If you format is not one of those, you can very likely use one of the all-purpose parsers supplied by the `taxa` package. You can read about those here:

https://github.com/ropensci/taxa#parsing-data


### Does this work with phyloseq?

Kind of. Metacoder and phyloseq use different data formats, but they both can store the same information. The `parse_phyloseq` converts from the `phyloseq` object to the `taxmap` object format that `metacoder` uses. The `as_phyloseq` function converts from `taxmap` to `phyloseq`, so you can use both on the same data.


### How do I subset a table for just one taxonomic rank (or other per-taxon data)?

This answer also applies to subsetting by any other per-taxon piece of information, such as taxon abundance or number of supertaxa.

Some functions in `metacoder` return per-taxon information, such as `calc_taxon_abund`.
All taxonomic ranks are represented in this type of output, so you if you see a row for "proteobacteria", "Bacteria" is probably also there (depending on your taxonomy). 
Often users want to subset these tables for rows corresponding to a given rank, such as all the species.
Lets get an example data set to play with:

```{r}
library(metacoder)
library(dplyr)

# Parse data for example
x = parse_tax_data(hmp_otus, class_cols = "lineage", class_sep = ";",
                   class_key = c(tax_rank = "taxon_rank", tax_name = "taxon_name"),
                   class_regex = "^(.+)__(.+)$")
                   
# Calculate the taxon abundance for each numeric column (i.e. sample)
x$data$tax_abund <- calc_taxon_abund(x, "tax_data", groups = hmp_samples$sex)

# Add taxon name to tax_abund to see filtering effects easier
x$data$tax_abund <- x$data$tax_abund %>%
  mutate(taxon = taxon_names(x)) %>%
  select(taxon_id, taxon, everything())

x
```

Say I want to subset "tax_abund" to just values for genera.
Typically, people try this first:

```{r}
filter_taxa(x, taxon_ranks == "g")
```

That will ruin you taxonomy since all the other taxa got filtered out, but as a side effect, it does filter that table as intended.
You can use `get_dataset()` to pull out the table without saving the resulting bad taxonomy:

```{r}
filter_taxa(x, taxon_ranks == "g") %>%
  get_dataset("tax_abund")
```

You might be tempted to use `filter_obs`, since you are trying to filter a specific table:

```{r}
filter_obs(x, "tax_abund", taxon_ranks == "g")
```

This does what we want, but only because "tax_abund" corresponds to the taxa one-to-one, so `taxon_names()` returns information that also applies to "tax_abund".
However, if you change the taxonomy  or "tax_abund" at all, this is no longer true and you can get errors like this:

```{r error=TRUE}
x %>%
  filter_taxa(n_obs > 10) %>%
  filter_obs("tax_abund", taxon_ranks == "g")
```

This happens because the taxonomy was filtered, so `taxon_ranks` changed, but rows were not removed from "tax_abund" since rows of deleted taxa were automatically reassigned to the deleted taxa's supertaxa.
This is desirable for data like the OTU table (you don't lose OTUs because a taxon is filtered), but makes what we are trying to do harder.
You can disable this behavior like so:

```{r}
x %>%
  filter_taxa(n_obs > 10, reassign_obs = FALSE) %>%
  filter_obs("tax_abund", taxon_ranks == "g")
```

This worked, but we also filtered the OTU table, which might be not be what we wanted to do. 
We can instead do the following to make that option apply to just "tax_abund":

```{r}
x %>%
  filter_taxa(n_obs > 10, reassign_obs = c(tax_abund = FALSE)) %>%
  filter_obs("tax_abund", taxon_ranks == "g")
```

So that works well. 
However, we are still relying on the taxa in the taxonomy to line up with the table, so using `filter_taxa` and not saving the whole object, as described above, is the most reliable strategy. 


### Error: cannot coerce type 'environment' to ...

Did you do this?

```{r error=TRUE}
library(metacoder)
ex_taxmap %>%
  filter_taxa(ex_taxmap, n_obs > 10)
```

Note the extra `ex_taxmap` in the call to `filter_taxa`.
Thats the same as this:

```{r error=TRUE}
filter_taxa(ex_taxmap, ex_taxmap, n_obs > 10)
```


### Can I only show some labels but keep the nodes in a heat tree?

Yes! Its even somewhat easy:

```{r eval=FALSE}
dont_print <- c("unknown", "unidentified")
heat_tree(
  ...,
  node_label = ifelse(taxon_names %in% dont_print, "", taxon_names),
  ...
)

```




### Is there a reccomended protocol or best practices guide? 

Not one made by us currently and we probably wont make one anytime soon.
This is because `metacoder` and `taxa` are meant to be data analysis *tools*, not the analysis itself.
Even the tutorials in this documentation are not meant to be used as is or considered a "best practice".
Best practices are a moving target, so we have focused on developing more abstract tools that should useful for many purposes.
We try to make [hammers and screwdrivers, not assembly lines](https://en.wikipedia.org/wiki/Unix_philosophy#Do_One_Thing_and_Do_It_Well).
Also, every dataset and study are different and something that might make sense for one might completely invalidate another.
**I highly reccomend against copying large sections of code in this documentation and using it for your analysis without understanding what the code is doing.**
On the other hand, if you do understand what an example analysis is doing, I welcome you to use it however you want!


### How do I cite metacoder?

How kind of you to ask!

Foster Z, Sharpton T and Grunwald N (2017). "Metacoder: An R package for visualization and manipulation of community taxonomic diversity data." PLOS Computational Biology, 13(2), pp. 1-15. doi: 10.1371/journal.pcbi.1005404


### How can I make a heat tree look the same each time?

Some of the layout options in `heat_tree` have a random component, so each time you run the function they will appear different.
If you have multiple figures, each using the same taxonomy, you might want them all to have the same layout so that they are easy to compare.
Or perhaps, you want a single tree to just always appear the same way for reproducibility. 
Most "random" things in computers are not actually random, but instead rely on [pseudorandom number generators](https://en.wikipedia.org/wiki/Pseudorandom_number_generator).
A function is given an initial "seed" (often the number of microseconds in the computers clock at the time by default) and returns some other apparently unrelated number.
That other number is then supplied to the function again and next time a random value is needed and so on.
The numbers may appear random, but a given seed will always result in the same series of numbers, and thus determine the result of any analysis with a random component.
In R, you can set the random seed whenever you want using `set.seed`.
For example, I can make a set of random numbers like so:

```{r}
rnorm(5)
```

If I run it again, it will be different:

```{r}
rnorm(5)
```

However, if I set a seed before each run, they will always be the same:

```{r}
set.seed(1)
rnorm(5)
set.seed(1)
rnorm(5)
```

You can do the same thing before calling `heat_tree` to make the layout always be the same for a given taxonomy and different seeds will produce different layouts.
Note how this layout has a random component:

```{r}
library(metacoder)
heat_tree(ex_taxmap, layout = "da")
heat_tree(ex_taxmap, layout = "da")
```

You can make it the same each time like so: 

```{r}
set.seed(1)
heat_tree(ex_taxmap, layout = "da")
set.seed(1)
heat_tree(ex_taxmap, layout = "da")
```

You can also use seeds to choose a specific layout you like since each seed will change the layout:

```{r}
set.seed(2)
heat_tree(ex_taxmap, layout = "da")
set.seed(2)
heat_tree(ex_taxmap, layout = "da")
```

