---
title: "Metacoder documentation"
author: "Zachary S. L. Foster and Niklaus J. Grünwald"
date: "`r Sys.Date()`"
bibliography: bibliography.bibtex
---

```{r home_setup, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
if (! is.null(current_input())) { # if knitr is being used
  read_chunk("settings.R")
} else {
  source("settings.R")
}
```

```{r rendering_settings, echo=FALSE, warning=FALSE, message=FALSE}
```


## Introduction

Metacoder is an R package for `r gloss$add("parsing")`, plotting, and manipulating large taxonomic data sets, like those generated from modern high-throughput sequencing, like metabarcoding (i.e. amplification metagenomics, 16S metagenomics, etc).
It provides a tree-based visualization called "heat trees" used to depict statistics for every taxon in a taxonomy using color and size.
It also provides various functions to do common tasks in microbiome bioinformatics on data in the `taxmap` format defined by the `taxa` package, such as:

* Summing read counts/abundance per taxon
* Converting counts to proportions and `r gloss$add("rarefaction")` of counts using `vegan`
* Comparing the abundance (or other characteristics) of groups of samples (e.g., experimental treatments) per taxon
* Combining data for groups of samples
* Simulated PCR, via EMBOSS primersearch, for testing primer specificity and coverage of taxonomic groups
* Converting common microbiome formats for data and reference databases into the objects defined by the `taxa` package.
* Converting to and from the `phyloseq` format and the `taxa` format

## Relationship with other packages

Many of these operations can be done using other packages like `phyloseq`, which also provides tools for diversity analysis.
The main strength of `metacoder` is that its functions use the flexible data types defined by `taxa`, which has powerful parsing and subsetting abilities that take into account the hierarchical relationship between taxa and user-defined data.
In general, `metacoder` and `taxa` are more of an abstracted tool kit, whereas `phyloseq` has more specialized functions for community diversity data, but they both can do similar things.
I encourage you to try both to see which fits your needs and style best.
You can also combine the two in a single analysis by converting between the two data types when needed.

## Installation

This project is available on CRAN and can be installed like so:

```{r eval = FALSE}
install.packages("metacoder")
```

You can also install the development version for the newest features, bugs, and bug fixes:

```{r eval = FALSE}
install.packages("devtools")
devtools::install_github("grunwaldlab/metacoder")
```

## Dependencies

The function that simulates PCR requires `primersearch` from the EMBOSS [@rice2000emboss] tool kit to be installed. 
This is not an R package, so it is not automatically installed.
Type `?primersearch` after installing and loading metacoder for installation instructions.


## Future development

Metacoder is under active development and many new features are planned.
Some improvements that are being explored include:

* Barcoding gap analysis and associated plotting functions
* A function to aid in retrieving appropriate sequence data from NCBI for *in silico* PCR from whole genome sequences
* Graphing of different node shapes in heat trees, possibly including pie graphs or [PhyloPics](http://phylopic.org/).
* Adding the ability to plot specific edge lengths in the heat trees so they can be used for phylogenetic trees.
* Adding more data import and export functions to make parsing and writing common formats easier. 

To see the details of what is being worked on, check out the [issues](https://github.com/grunwaldlab/metacoder/issues) tab of the Metacoder [Github site](https://github.com/grunwaldlab).


## Aknowledgements

Metacoder's major dependencies are `taxa`, `taxize`, `vegan`, `igraph`, `dplyr`, and `ggplot2`.


## Feedback and contributions

We would like to hear about users' thoughts on the package and any errors they run into.
Please report errors, questions or suggestions on the [issues](https://github.com/grunwaldlab/metacoder/issues) tab of the Metacoder [Github site](https://github.com/grunwaldlab).
We also welcome contributions via a Github [pull request](https://help.github.com/articles/using-pull-requests/).
You can also talk with us using our [Google groups](https://groups.google.com/forum/#!forum/metacoder-discussions) site or the comments section below. 


## References