---
title: Graphing the RDP database
---


```{r greengenes_setup, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE}
knitr::read_chunk("settings.R")
```

```{r rendering_settings, echo = FALSE, warning=FALSE, message=FALSE, cache=FALSE}
```

## Requirements 

**NOTE:** This analysis requires at least 10Gb of RAM to run. 

## Parameters

```{r io_settings, cache=FALSE}
```

```{r database_comparison_settings, cache=FALSE}
```


## Parse database

The code below parses and subsets the entire RDP non-redundant reference database.
The object made is quite large.

```{r rdp_load}
library(metacoder)
file_path <- file.path(input_folder, "rdp_current_Bacteria_unaligned.fa")
system.time(rdp <- extract_taxonomy(seqinr::read.fasta(file_path, as.string = TRUE),
                                    regex = "\\tLineage(.*)",
                                    key = c("class"),
                                    class_regex = "[;=](.+?);(.+?)",
                                    class_key = c("name", rdp_rank = "taxon_info")))
print(rdp)
```

## Subset

```{r rdp_subset}
if (! is.null(min_seq_count)) {
  system.time(rdp <- filter_taxa(rdp, n_obs >= min_seq_count))
}
if (just_bacteria) {
  system.time(rdp <- filter_taxa(rdp, name == "Bacteria", subtaxa = TRUE))
}
if (! is.null(max_taxonomy_depth)) {
  system.time(rdp <- filter_taxa(rdp, n_supertaxa <= max_taxonomy_depth))
}
print(rdp)
```


## Remove chloroplast sequences

These are not bacterial and will bias the *in silico* PCR results.

```{r rdp_rm_chloro}
system.time(rdp <- filter_taxa(rdp, name == "Cyanobacteria/Chloroplast", subtaxa = TRUE, invert = TRUE))
print(rdp)
```


## Plot whole database

Although RDP is such a large database (`r nrow(rdp$taxa_data)` taxa) that graphing everything can be a bit overwhelming, it gives an intuitive feel for the complexity of the database:

```{r rdp_plot_all}
system.time(rdp_plot_all <- heat_tree(rdp,
                                      node_size = n_obs,
                                      node_color = n_obs,
                                      node_size_range = size_range * 2,
                                      edge_size_range = size_range,
                                      node_size_interval = all_size_interval,
                                      edge_size_interval = all_size_interval,
                                      node_color_interval = all_size_interval,
                                      edge_color_interval = all_size_interval,
                                      node_label = name,
                                      node_label_size_range = label_size_range,
                                      node_label_max = label_max,
                                      node_color_axis_label = "Sequence count",
                                      make_legend = TRUE,
                                      output_file = file.path(output_folder, paste0("rdp--all", output_format))))
print(rdp_plot_all)
```


## PCR

```{r rdp_length_filter}
if (! is.null(min_seq_length)) {
  system.time(rdp <- filter_obs(rdp, nchar(sequence) >= min_seq_length, unobserved = FALSE))
}
print(rdp)
```

```{r rdp_pcr}
system.time(rdp_pcr <- primersearch(rdp,
                                    forward = forward_primer,
                                    reverse = reverse_primer,
                                    mismatch = max_mismatch))
```

```{r rdp_plot_pcr_all}
system.time(rdp_plot_pcr_all <- heat_tree(rdp_pcr,
                                          node_size = n_obs,
                                          node_label = name,
                                          node_color = prop_amplified,
                                          node_color_range =  pcr_success_color_scale,
                                          node_color_trans = "linear",
                                          node_label_size_range = label_size_range,
                                          node_label_max = 150,
                                          edge_color_interval = c(0, 1),
                                          node_color_interval = c(0, 1),
                                          node_color_axis_label = "Proportion PCR success",
                                          node_size_axis_label = "Sequence count",
                                          output_file = file.path(output_folder, paste0("rdp--pcr_all", output_format))))
print(rdp_plot_pcr_all)
```

```{r, rdp_plot_pcr_fail}
system.time(rdp_plot_pcr_fail <- rdp_pcr %>%
              filter_taxa(prop_amplified < pcr_success_cutoff, supertaxa = TRUE) %>%
              heat_tree(node_size = n_obs - count_amplified,
                        node_label = name,
                        node_color = prop_amplified,
                        node_size_range = size_range * 2,
                        edge_size_range = size_range,
                        node_size_interval = pcr_size_interval,
                        edge_size_interval = pcr_size_interval,
                        node_color_range =  pcr_success_color_scale,
                        node_color_trans = "linear",
                        node_color_interval = c(0, 1),
                        edge_color_interval = c(0, 1),
                        node_label_size_range = label_size_range,
                        node_label_max = 1000,
                        node_color_axis_label = "Proportion PCR success",
                        node_size_axis_label = "Sequences not amplified",
                        make_legend = TRUE,
                        output_file = file.path(output_folder, paste0("rdp--pcr_fail", output_format))))
print(rdp_plot_pcr_fail)
```


## Save outputs for composite figure

Some results from this file will be combined with other to make a composite figure.
Below, the needed objects are saved so that they can be loaded by another Rmd file.

```{r rdp_save}
save(file = file.path(output_folder,"rdp_data.RData"),
     rdp_plot_all, rdp_plot_pcr_fail)
```

