---
title: Graphing the SILVA database
---

```{r greengenes_setup, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE}
knitr::read_chunk("settings.R")
```

```{r rendering_settings, echo = FALSE, warning=FALSE, message=FALSE, cache=FALSE}
```

## Requirements 

**NOTE:** This analysis requires at least 10Gb of RAM to run. 

## Parameters

```{r io_settings, cache=FALSE}
```

```{r database_comparison_settings, cache=FALSE}
```


## Parse database

The code below parses and subsets the entire SILVA non-redundant reference database.
The object made is quite large.

```{r load_silva}
library(metacoder)
file_path <- file.path(input_folder, "SILVA_123.1_SSURef_Nr99_tax_silva.fasta")
system.time(silva <- extract_taxonomy(seqinr::read.fasta(file_path, as.string = TRUE), 
                                      regex = "^(.*?) (.*)$",
                                      key = c(id = "obs_info", "class"),
                                      class_sep = ";"))
print(silva)
```

## Subset

```{r subset_silva}
if (! is.null(max_taxonomy_depth)) {
  system.time(silva <- filter_taxa(silva, n_supertaxa <= max_taxonomy_depth))
}
if (! is.null(min_seq_count)) {
  system.time(silva <- filter_taxa(silva, n_obs >= min_seq_count))
}
if (just_bacteria) {
  system.time(silva <- filter_taxa(silva, name == "Bacteria", subtaxa = TRUE))
}
if (! is.null(min_seq_length)) {
  system.time(silva <- filter_obs(silva, nchar(sequence) >= min_seq_length, unobserved = FALSE))
}
print(silva)
```


## Remove chloroplast sequences

These are not bacterial and will bias the *in silico* PCR results.

```{r greengenes_rm_chloro}
system.time(silva <- filter_taxa(silva, name == "Chloroplast", subtaxa = TRUE, invert = TRUE))
print(silva)
```


## Plot whole database

Although SILVA is such a large database (`r nrow(silva$taxa_data)` taxa) that graphing everything can be a bit overwhelming, it gives an intuitive feel for the complexity of the database:

```{r silva_plot_all}
system.time(silva_plot_all <- plot(silva,
                                   node_size = n_obs,
                                   node_color = n_obs,
                                   node_size_range = c(0.0005, 0.012) * 2,
                                   edge_size_range = c(0.0005, 0.012),
                                   node_size_interval = c(1, 3000000),
                                   edge_size_interval = c(1, 3000000),
                                   node_color_interval = c(1, 3000000),
                                   edge_color_interval = c(1, 3000000),
                                   node_label = name,
                                   node_label_max = 150,
                                   node_color_axis_label = "Sequence count",
                                   make_legend = TRUE,
                                   output_file = file.path(output_folder, paste0("silva--all", output_format))))
print(silva_plot_all)
```


## PCR

```{r silva_pcr}
# Replace all u with t so in silico PCR works
system.time(silva$obs_data$sequence <- gsub(pattern = "u", replacement = "t", silva$obs_data$sequence))
# in silico PCR
system.time(silva_pcr <- primersearch(silva,
                                      forward = forward_primer,
                                      reverse = reverse_primer,
                                      mismatch = max_mismatch))
```

```{r silva_plot_pcr_all}
system.time(silva_plot_pcr_all <- plot(silva_pcr,
                                       node_size = n_obs,
                                       node_label = name,
                                       node_color = prop_amplified,
                                       node_color_range =  pcr_success_color_scale,
                                       node_color_trans = "linear",
                                       node_label_max = 150,
                                       node_color_axis_label = "Proportion PCR success",
                                       node_size_axis_label = "Sequence count",
                                       output_file = file.path(output_folder, paste0("silva--pcr_all", output_format))))
print(silva_plot_pcr_all)
```

```{r, silva_plot_pcr_fail}
system.time(silva_plot_pcr_fail <- silva_pcr %>%
              filter_taxa(prop_amplified < pcr_success_cutoff, supertaxa = TRUE) %>%
              plot(node_size = n_obs - count_amplified,
                   node_label = name,
                   node_color = prop_amplified,
#                    node_size_range = c(0.0005, 0.012) * 2,
#                    edge_size_range = c(0.0005, 0.012),
#                    node_size_interval = c(1, 1000000),
#                    edge_size_interval = c(1, 1000000),
                   edge_color_interval = c(0, 1),
                   node_color_range =  pcr_success_color_scale,
                   node_color_trans = "linear",
                   node_color_interval = c(0, 1),
                   node_label_size_range = c(0.008, 0.015),
                   node_label_max = 500,
                   node_color_axis_label = "Proportion PCR success",
                   node_size_axis_label = "Sequences not amplified",
                   make_legend = TRUE,
                   output_file = file.path(output_folder, paste0("silva--pcr_fail", output_format))))
print(silva_plot_pcr_fail)
```


## Save outputs for composite figure

Some results from this file will be combined with other to make a composite figure.
Below, the needed objects are saved so that they can be loaded by another Rmd file.

```{r silva_save}
save(file = file.path(output_folder, "silva_data.RData"),
     silva_plot_all, silva_plot_pcr_fail)
```

