---
title: Graphing voting geography
---

```{r voting_setup, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE}
knitr::read_chunk("settings.R")
```

```{r rendering_settings, echo = FALSE, warning=FALSE, message=FALSE, cache=FALSE}
```

```{r io_settings, echo = FALSE, warning=FALSE, message=FALSE, cache=FALSE}
```

## Read in data

```{r}
file_path <- file.path(input_folder, "primary_results.csv")
data <- tibble::as_tibble(read.table(file_path, sep = ",", header = TRUE, quote = "", stringsAsFactors = FALSE))
```

## Add region column

```{r}
divisions <- stack(list("New England" = c("Connecticut", "Maine", "Massachusetts",
                                          "New Hampshire", "Rhode Island", "Vermont"),
                        "Mid-Atlantic" = c("New Jersey", "New York", "Pennsylvania"),
                        "NE Central" = c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin"),
                        "NW Central" = c("Iowa", "Kansas", "Minnesota", "Missouri",
                                                 "Nebraska", "North Dakota", "South Dakota"),
                        "South Atlantic" = c("Delaware", "Florida", "Georgia", "Maryland",
                                             "North Carolina", "South Carolina", "Virginia",
                                             "Washington D.C.", "West Virginia"),
                        "SE Central" = c("Alabama", "Kentucky", "Mississippi", "Tennessee"),
                        "SW Central" = c("Arkansas", "Louisiana", "Oklahoma", "Texas"),
                        "Mountain" = c("Arizona", "Colorado", "Idaho", "Montana", "Nevada",
                                       "New Mexico", "Utah", "Wyoming"),
                        "Pacific" = c("Alaska", "California", "Hawaii", "Oregon", "Washington")))
data$division <- divisions$ind[match(data$state, divisions$values)]
regions <- stack(list("Northeast" = c("New England", "Mid-Atlantic"),
                      "Midwest" = c("NE Central", "NW Central"),
                      "South" = c("South Atlantic", "SE Central", "SW Central"),
                      "West" = c("Mountain", "Pacific")))
data$region <- regions$ind[match(data$division, regions$values)]
data[] <- lapply(data, as.character)
```


## Create and parse classifications

The code below creates a single column in the dataset that contains all of the levels of the geographic hierarchy for each location. 
It is then parsed using `parse_taxonomy_table` so that the other columns are preserved in the `taxmap` object.

```{r}
library(metacoder)
data$classification <- paste(sep = "|", "USA", data$region, data$division, data$state, data$county)
voting_data <- parse_taxonomy_table(input = data, taxon_col = c("class" = -1), other_col_type = "obs_info", class_sep = "\\|")
```

## Get canidate vote counts

Instead of adding a static column for votes for Bernie / Hillary, I will add column-generating functions so that subsets of the dataset automitically updated vote counts.

```{r}
voting_data$taxon_funcs <- c(voting_data$taxon_funcs, 
                             total_votes = function(voting_data, subset) {
                               sapply(obs(voting_data, subset),
                                      function(i) sum(voting_data$obs_data$votes[i]))
                             },
                             clinton_votes = function(voting_data, subset) {
                               sapply(obs(voting_data, subset),
                                      function(i) {
                                        subset <- voting_data$obs_data[i, ]
                                        sum(subset$votes[subset$candidate == "Hillary Clinton"])
                                      })
                             },
                             sanders_votes = function(voting_data, subset) {
                               sapply(obs(voting_data, subset),
                                      function(i) {
                                        subset <- voting_data$obs_data[i, ]
                                        sum(subset$votes[subset$candidate == "Bernie Sanders"])
                                      })
                             })
```

## Plotting results

```{r}
voting_data %>%
  filter_taxa(total_votes >= 100) %>%
  heat_tree(node_size = total_votes,
            node_size_range = c(0.0002, 0.06),
            node_color = (clinton_votes - sanders_votes) / total_votes * 100,
            node_label = ifelse(n_supertaxa <= 3, name, NA),
            edge_label = ifelse(total_votes > 100000 & n_supertaxa == 4, name, NA),
            node_color_range = c("cyan", "lightgray", "magenta"),
            node_color_interval = c(-50, 50),
            edge_color_range = c("cyan", "lightgray", "magenta"),
            edge_color_interval = c(-50, 50),
            margin_size = c(0.001, 0.001),
            node_color_axis_label = "Clinton                                Sanders",
            node_size_axis_label = "Total votes",
            node_label_max = 500,
            edge_label_max = 500, 
            overlap_avoidance = 10,
            # layout = "da", initial_layout = "re",
            output_file = file.path(output_folder,
                                    paste0("voting",
                                           output_format)))
```

