---
title: "Tara oceans example"
---


```{r tara_setup, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE}
knitr::read_chunk("settings.R")
```

```{r rendering_settings, echo = FALSE, warning=FALSE, message=FALSE, cache=FALSE}
```

## Requirements 

**NOTE:** This analysis requires at least 10Gb of RAM to run. 

## Parameters

```{r io_settings, cache=FALSE}
```

```{r tara_parameters}
```


## Parsing taxonomic data

The data set at the below URL was downloaded and uncompressed:

http://taraoceans.sb-roscoff.fr/EukDiv/data/Database_W5_OTU_occurences.tsv.zip

```{r tara_parse, warning=FALSE, message=FALSE}
library(metacoder)
data <- parse_taxonomy_table(file.path(input_folder, "Database_W5_OTU_occurences.tsv"), 
                             taxon_col = c("class" = -9), class_sep = "\\|")
```


## Getting sample data

The sample data was downloaded from the URL below:

http://taraoceans.sb-roscoff.fr/EukDiv/data/Database_W1_Sample_parameters.xls

```{r tara_parse_sample}
sample_data <- readxl::read_excel(file.path(input_folder, "Database_W1_Sample_parameters.xls"))
```


## Caluculate read abundance per taxon

```{r tara_calculate_abundance}
sample_columns <- sample_data[["PANGAEA ACCESSION NUMBER"]]
data <- mutate_taxa(data,
                    read_counts = vapply(obs(data),
                                         function(x) sum(data$obs_data$totab[x]), numeric(1)))
```


## Plot read and OTU abundance

```{r tara_plot_all}
seed = 9 #9, 10, 12 is good
set.seed(seed)
taxa_patterns_to_hide <- paste0("^", c("[X]+", "X\\+sp\\.", "NA", "root", "\\*", "sp\\.", "sp"), "$")
taxa_patterns_to_remove <- paste0("^", c("X\\+sp\\.", "NA", "root", "\\*", "sp\\.", "sp"), "$")
background_color <- "#ccfff7"
data %>%
  filter_taxa(! Reduce(`|`, lapply(taxa_patterns_to_remove, grepl, x = name))) %>%
  filter_taxa(read_counts >= 100) %>%
  filter_taxa(!(taxon_ids %in% subtaxa(data, name == "Eukaryota", simplify = TRUE, include_input = TRUE) & n_supertaxa <= 1)) %>%
  heat_tree(title = "Plankton diversity in the sunlit ocean",
            title_size = 0.03,
            node_color_axis_label = "Number of reads (Abundance)",
            node_size_axis_label = "Number of species (OTUs)",
            node_size = n_obs,
            node_size_range = c(0.0012, NA),
            node_color = read_counts,
            node_color_range = c("grey", "#80cdc1", "#018571", "#dfc27d", "#a6611a"),
            node_color_trans = "log10",
            node_label = ifelse(grepl(pattern = "^[a-zA-z\\-]{1,25}$", name) &
                                  ! Reduce(`|`, lapply(taxa_patterns_to_hide, grepl, x = name)) &
                                  read_counts > 10000,
                                name, NA),
            node_label_color = "#000000",
            node_label_color_trans = "area",
            node_label_size = (n_obs / (n_supertaxa + 1)) ^ 0.5,
            node_label_size_trans = "area",
            node_label_size_range = c(0.001, NA),
            node_label_max = 1000,
            tree_label = name,
            tree_label_color = "#00806c",
            tree_label_max = 100,
            initial_layout = "re", layout = "da",
            overlap_avoidance = .65,
            background_color = background_color,
            maxiter = 50, fineiter = 50,
            margin_size = c(0.001, 0.001),
            output_file = file.path(output_folder, paste0("sup_figure_1--tara_all_plankton_diversity",  output_format)))
```


## Plot propotion of OTUs identified 

For each taxon we will identify the proportion of OTUs with less than a 90% match to the most similar reference sequence to approximate classification certainty. 

```{r tara_calc_identified}
data <- mutate_obs(data, pid = as.numeric(pid))
data <- mutate_taxa(data, mean_pid = vapply(obs(data),
                                            function(x) mean(data$obs_data$pid[x], na.rm = TRUE), numeric(1)))
# Percentage of OTUs with less than 90% idententiy
data <- mutate_taxa(data, percent_known = vapply(obs(data),
                                                 function(x) sum(data$obs_data$pid[x] >= 90, na.rm = TRUE) / length(x) * 100, numeric(1)))
```

```{r tara_plot_all_identified}
seed = 13
set.seed(seed)
taxa_patterns_to_hide <- paste0("^", c("[X]+", "X\\+sp\\.", "NA", "root", "\\*", "sp\\.", "sp"), "$")
taxa_patterns_to_remove <- paste0("^", c("[X]+", "X\\+sp\\.", "NA", "root", "\\*", "sp\\.", "sp"), "$")
data %>%
  filter_taxa(! Reduce(`|`, lapply(taxa_patterns_to_remove, grepl, x = name))) %>%
  filter_taxa(read_counts >= 100) %>%
  filter_taxa(!(taxon_ids %in% subtaxa(data, name == "Eukaryota", simplify = TRUE, include_input = TRUE) & n_supertaxa <= 1)) %>%
  filter_taxa(n_subtaxa < 30 & n_supertaxa == 0, subtaxa = TRUE, invert = TRUE) %>%
  heat_tree(
    # title = "Poportion of OTUs not well identified",
    title_size = 0.03,
    node_color_axis_label = "Percent of species (OTUs) identified",
    node_size_axis_label = "Number of species (OTUs)",
    node_size = n_obs,
    node_size_range = c(0.0012, NA),
    node_color = percent_known,
    node_color_range = c("red", "orange", "yellow", "green", "cyan"),
    node_color_trans = "linear",
    node_color_interval = c(0, 100),
    # node_label = ifelse(grepl(pattern = "^[a-zA-z\\-]{1,25}$", name) &
    #                       ! Reduce(`|`, lapply(taxa_patterns_to_hide, grepl, x = name)),
    #                     name, NA),
    node_label_size = (n_obs / (n_supertaxa + 1)) ^ 0.5,
    node_label_size_trans = "area",
    node_label_size_range = c(0.001, NA),
    node_label_max = 1000,
    tree_label = name,
    tree_label_max = 100,
    tree_label_size_range = c(0.008, 0.035),
    initial_layout = "re", layout = "da",
    overlap_avoidance = .65,
    maxiter = 50, fineiter = 50,
    margin_size = c(0.001, 0.001),
    make_legend = FALSE,
    output_file = file.path(output_folder, paste0("sup_figure_2--tara_proportion_identified",  output_format)))
```

### Just Metazoa

```{r tara_plot_metazoa_identified}
set.seed(64)
taxa_patterns_to_hide <- paste0("^", c("[X]+", "X\\+sp\\.", "NA", "root", "\\*", "sp\\.", "sp"), "$")
taxa_patterns_to_remove <- paste0("^", c("[X]+", "X\\+sp\\.", "NA", "root", "\\*", "sp\\.", "sp"), "$")
data %>%
  filter_taxa(! Reduce(`|`, lapply(taxa_patterns_to_remove, grepl, x = name))) %>%
  filter_taxa(read_counts >= 200) %>%
  filter_taxa(name == "Metazoa", subtaxa = TRUE) %>%
  heat_tree(node_color_axis_label = "Percent of OTUs identified",
            edge_color_axis_label = "Percent of OTUs identified",
            edge_size = read_counts,
            edge_size_range = c(0.001, 0.01),
            edge_color = percent_known,
            edge_color_range = c("red", "orange", "yellow", "greenyellow", "green"),
            edge_color_trans = "linear",
            node_size_axis_label = "Number of OTUs",
            edge_size_axis_label = "Number of reads",
            node_size = n_obs,
            node_size_range = c(0.005, 0.03),
            node_color = percent_known,
            node_color_range = c("red", "orange", "yellow", "greenyellow", "green"),
            node_color_trans = "linear",
            node_color_interval = c(0, 100),
            node_label = ifelse(grepl(pattern = "^[a-zA-z\\-]{1,25}$", name) &
                                  ! Reduce(`|`, lapply(taxa_patterns_to_hide, grepl, x = name)),
                                name, NA),
            node_label_size = (n_obs / (n_supertaxa + 1)) ^ 0.5,
            node_label_size_range = c(0.008, 0.012),
            node_label_size_trans = "area",
            node_label_max = 1000,
            initial_layout = "fr", layout = "da",
            overlap_avoidance = .8,
            maxiter = 50, fineiter = 50,
            weight.node.edge.dist = 15,
            output_file = file.path(output_folder,
                                    paste0("figure_1--tara_metazoa--seed_",
                                           seed, output_format)))
```
